Beliefs about the development of mental life

### 2024-12-19 paper/paper.Rmd
massive overhaul

Indeed, caring for a young child involves making countless decisions under uncertainty: Why is my child crying, smiling, shouting, or biting? What do they want, and what do they need? Can they understand what I am saying? Do they love me? As in other domains of reasoning [@gerstenberg.tenenbaum2017; @gopnik.wellman1994; @wellman.gelman1992], people likely draw on folk beliefs and intuitive theories about how the mind develops to predict, explain, and respond to children’s behavior and development, with consequences for the well-being of children in their care.

These folk beliefs and intuitive theories are, to date, poorly understood by social scientists. A handful of recent studies have examined the extent to which cognitive abilities, knowledge, and emotions are thought to be innate vs. learned, with special attention to comparing laypeople’s beliefs about innateness to scientific evidence for or against innateness and universality [@wang.feigenson2019; @berent.etal2019; @berent.etal2020; see @berent2021 for discussion]. These studies offer a fascinating glimpse of potential gaps, misconceptions, and biases in laypeople’s beliefs about the human mind, but they leave open many questions about the nature of the folk beliefs and intuitive theories themselves. What is the cognitive architecture underlying laypeople’s beliefs about the origins of the human mind? How do laypeople "carve up" the various sensations, experiences, and abilities that make up human mental life, and—most crucially from the perspective of caregiving—how do laypeople understand changes in mental life across early development?

In this paper, we examine how US adults conceptualize the developing minds of infants and children by drawing on techniques developed in studies of mind perception, in which researchers use large datasets and modern statistical methods to reveal conceptual representations of "the mind," broadly construed. In a groundbreaking study, Gray et al. asked participants to compare the mental capacities of a range of humans, animals, technologies, and other entities, and argued that participants’ judgments revealed a fundamental distinction between "agency" (e.g., capacities for self-control, moral reasoning, memory, and planning) and "experience" (e.g., capacities for hunger, fear, pain, and pleasure) [@gray.etal2007]. More recently, Weisman et al. used a similar empirical approach to argue that people’s conceptions of mental life are anchored by three fundamental components: "body" (physiological sensations such as hunger and pain), "heart" (social-emotional abilities such as embarrassment and pride), and "mind" (perceptual-cognitive capacities such as vision and memory [@weisman.etal2017a; see @malle2019 for a similar three-factor solution]. Such findings have inspired subsequent work on topics as wide-ranging as moral reasoning, dehumanization and stereotyping, beliefs about the afterlife, and human-robot interactions.

@@

Yet infants and young children are far from unusual social partners. Caregiving relationships are at the core of human existence; for many people, infants and young children are among the most frequent and highly valued minds that we encounter. In the current studies we set out to shed new light on this important and understudied aspect of mind perception: how laypeople reason about the development of human mental life. In three studies, we assessed US adults’ folk beliefs about children’s developing abilities for physiological sensation, perception, cognition, emotion, self-regulation, social interaction, and other aspects of "mental life," broadly construed. We focused on development between birth and age five years—a time of particularly dramatic maturation in perceptual acuity, emotional experience, cognitive ability, executive function, social engagement, and many other domains.

Methods, inclusion/exclusion criteria, and analyses marked as "preregistered" were preregistered on the Open Science Framework (OSF) website: (anonymized links for peer review: Study 1: https://osf.io/e6ajh/?view_only=f2539dbce3664172907aa7a381f62d34; Study 2: https://osf.io/j72dg/?view_only=7e4c01fec32f40a3b13c0ebc2b7ced18; Study 3: https://osf.io/xh8ce/?view_only=56649aff37aa4f1db2dc06b2e69576f4). All studies were approved by [IRB information removed to preserve anonymity]. For extended descriptions of methods and results, see Supplemental Materials.

@@

We designed Study 1 to explore the correlational structure of participants’ responses, as a window into the underlying conceptual structure that might support their reasoning about the development of mental life. Our primary analysis was a preregistered exploratory factor analysis (EFA) of participants’ capacity ratings. This analysis allowed us to examine which capacities tend to "hang together" in participants’ assessments of the mental lives of infants and young children. For example, when a participant indicated that newborns are highly capable of feeling scared, what other capacities did they tend to attribute to newborns? Following previous work on mind perception, we argue that the suites of capacities revealed by applying dimensionality reduction (in our case, EFA) to this covariance structure offer a meaningful approximation of the latent conceptual structure underlying participants’ reasoning about the developing human mind.

@@

It is worth noting that a handful of items loaded relatively strongly on two factors (loadings $\geq$ 0.40), suggesting that these factors are not entirely distinct. In our view, these cross-loadings were generally sensible. The item "feeling distressed", for instance, loaded roughly equally strongly on both bodily sensations and negative affect—perhaps reflecting the idea that "distress" might be experienced both as a physical sensation (similar to pain) and an affective state (similar to frustration). Likewise, "seeing" loaded equally strongly on both _bodily sensations_ and _social connection_—perhaps reflecting an understanding of vision as both a basic sensory ability (closely tied to a particular body part) and a critical part of many social interactions (especially stereotypical interactions with infants, which often hinge on making eye contact, imitating facial expressions, playing peekaboo, etc.).

As a check on our choice of labels for these factors, we turned to OpenAI’s Generative Pre-trained Transformer 3 (GPT-3) [@brown.etal2020], an autoregressive transformer-based language prediction model trained on an enormous corpus of English-language text drawn primarily from “crawls” of the Internet as well as English-language novels and the open-source encyclopedia Wikipedia. GPT-3 is capable of generating remarkably human-like text in response to questions and other prompts. GPT-3, like other generative language models, was developed under the mandate of predicting the continuation of natural language sequences: Most concretely, then, its “responses” should be understood to reflect a sequence of words that is likely to follow the prompt it has been provided. In a more abstract sense, however, the content of these responses could be understood to represent the “common knowledge” embedded in the training data [for extended arguments relevant to this proposal, see @bhatia.richie2021; @lake.murphy2021; @petroni.etal2019]. For example, if prompted with the questions “What do you get when you mix red and yellow?” or “What’s a good name for a vegetarian restaurant?,” GPT-3 will generate text like “When you mix red and yellow, you get orange” or “The Green Plate” (respectively). In continuing the language sequence, the model also provides novel content that reflects actual human knowledge, biases, associations, and intuitions.

Using the openly available GPT-3 “playground” (https://beta.openai.com/playground), we provided a variety of prompts to the model to assess its “intuitions” about how to label the four categories of mental capacities suggested by the factor solution just described. One representative prompt read, “Name these four categories of mental capacities: Category 1 is […]. Category 2 is […],” and so on, with the ellipses replaced by lists of all the items for which each factor (“Category”) was the dominant factor. The response generated by GPT-3 in this case was as follows: “Category 1: Executive functioning skills. Category 2: Social and emotional skills. Category 3: Physical needs and sensations. Category 4: Negative emotions.” In nearly every variation of our prompts, GPT-3’s labels for the factor we have labeled cognition and control included the phrase “executive function”; its labels for the factor we have labeled social connection always nearly always included the word “social” or “social-emotional”; its labels for the factor we have labeled bodily sensation nearly always included the words “sensations” (and often “physical needs” or “basic needs”); and its labels for the factor we have labeled negative affect nearly always included the phrase “negative emotion.” (See Supplemental Materials for the full text of these prompts and GPT-3’s responses.) The similarity. between our original labels (i.e. cognition and control, social connection, bodily sensations, and negative affect) and those generated by GPT-3 provide some evidence that the labels used are consistent with common knowledge and intuitions reflected in the vast swath of English language text that constituted the training data for this model, and by extension with the common knowledge and intuitions shared by many lay people in the English-speaking world. Nonetheless, we caution the reader to treat these labels as interpretations of empirical results, rather than empirical results themselves.

@@

One limitation of this fixed-order design is that it makes it difficult to assess the stability of participants’ assessments of the fine-grained differences between specific points in development. See Supplemental Materials for a comparison of capacity ratings and summary scores by domain at the three target ages that were included in all of the current studies (i.e., newborns, 9-month-olds, and 5-year-olds), which provides some sense of the stability of participants’ representations of these particular target ages.

@@

A preregistered EFA yielded four factors very similar to the four factors from Study 1, with all capacity items loading strongly on the factors that they were selected to represent and less strongly on other factors. (See Supplemental Materials for this 4-factor solution, which was suggested by both parallel analysis and Weisman et al.’s (2017) factor retention criteria, as well as the 8-factor solution suggested by minimizing BIC). Minimally, this suggests that our selection of mental capacities for this study captured the essential meaning of the four factors revealed in Study 1 (including the factor we have labeled _negative affect_: In Study 2, this factor accounted for fully `r round_x(s2_results$efa$Vaccounted["Proportion Explained", "MR1"]*100, 0)`% of the shared variance and had strong factor loadings (≥ 0.60) for all 5 of the items selected to represent it, none of which loaded strongly on other factors). Beyond this, we consider these results to be a somewhat independent conceptual replication of Study 1 results and validation of our interpretation of these factors. After all, given that Study 2 featured only 20 of the 60 capacities featured in Study 1, it certainly could have been the case that this analysis would suggest retaining fewer factors (as often occurs with fewer variables). Conversely, given that Study 2 featured 13 within-subjects observations for each capacity compared to the 3 in Study 1, this analysis could have surfaced more than four factors (as often occurs with more observations). Finally, if we had misinterpreted the fundamental semantic content that distinguished the factors in Study 1 and selected inappropriate capacities to represent these factors, the Study 2 analysis could have identified more capacities with cross-loadings. Instead, the results of this analysis offer some validation of the four factors we have called _bodily sensation_, _negative affect_, _social connection_, and _cognition and control_.

@@

```{r s2 gam results, include = F}
s2_gam_ptab <- summary(s2_results$gam$gam)$p.table
s2_gam_stab <- summary(s2_results$gam$gam)$s.table
```

Regressing participants' item-level responses onto target age, domain, and interactions between them via a multilevel generalized additive model confirmed both of our preregistered hypotheses: On average, participants rated older children as more capable, but perceptions of the development of children’s capacities differed dramatically across domains. On one extreme, participants attributed very little capacity for _cognition and control_ to newborns (median score for newborns: `r s2_cap_rat_newborns_ss %>% filter(factor == "Cognition and control") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Cognition and control") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Cognition and control") %>% pull(sd) %>% round_x(., 2)`), but reported that such capacities increase dramatically over the target age range (average smoothing term, analogous to a slope in linear regression: $F$ = `r round_x(s2_gam_stab["s(target_year):factorCognition and control", "F"], 2)`, $p$ `r ifelse(s2_gam_stab["s(target_year):factorCognition and control", "p-value"] < 0.001, "< 0.001", paste("=", round_x(s2_gam_stab["s(target_year):factorCognition and control", "p"], 3)))`, $edf$ = `r round_x(s2_gam_stab["s(target_year):factorCognition and control", "edf"], 2)`). Compared to these attributions of _cognition and control_, participants attributed substantially greater capacities to newborns in the domains that we have labeled _social connection_ (median score for newborns: `r s2_cap_rat_newborns_ss %>% filter(factor == "Social connection") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Social connection") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Social connection") %>% pull(sd) %>% round_x(., 2)`; increase in log-odds relative to _cognition and control_: $\beta$ = `r round_x(s2_gam_ptab["factorsoc_cog", "Estimate"], 2)`, $p$ `r ifelse(s2_gam_ptab["factorsoc_cog", "Pr(>|t|)"] < 0.001, "< 0.001", paste("=", round_x(s2_gam_ptab["factorsoc_cog", "Pr(>|t|)"], 3)))`) and _negative affect_ (median score for newborns: `r s2_cap_rat_newborns_ss %>% filter(factor == "Negative affect") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Negative affect") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Negative affect") %>% pull(sd) %>% round_x(., 2)`; increase in log-odds: $\beta$ = `r round_x(s2_gam_ptab["factorneg_cog", "Estimate"], 2)`, $p$ `r ifelse(s2_gam_ptab["factorneg_cog", "Pr(>|t|)"] < 0.001, "< 0.001", paste("=", round_x(s2_gam_ptab["factorneg_cog", "Pr(>|t|)"], 3)))`) and, accordingly, reported smaller increases over the target age range in both domains (average smoothing terms for _social connection_: $F$ = `r round_x(s2_gam_stab["s(target_year):factorSocial connection", "F"], 2)`, $p$ `r ifelse(s2_gam_stab["s(target_year):factorSocial connection", "p-value"] < 0.001, "< 0.001", paste("=",  round_x(s2_gam_stab["s(target_year):factorCognition and control", "p"], 3)))`, $edf$ = `r round_x(s2_gam_stab["s(target_year):factorSocial connection", "edf"], 2)`; for _negative affect_: $F$ = `r round_x(s2_gam_stab["s(target_year):factorNegative affect", "F"], 2)`, $p$ `r ifelse(s2_gam_stab["s(target_year):factorNegative affect", "p-value"] < 0.001, "< 0.001", paste("=", round_x(s2_gam_stab["s(target_year):factorNegative affect", "p"], 3)))`, $edf$ = `r round_x(s2_gam_stab["s(target_year):factorCognition and control", "edf"], 2)`). At the other extreme, in the domain of _bodily sensation_ participants attributed nearly maximal capacity to newborns (median score for newborns: `r s2_cap_rat_newborns_ss %>% filter(factor == "Bodily sensation") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Bodily sensation") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Bodily sensation") %>% pull(sd) %>% round_x(., 2)`; increase in log-odds relative to _cognition and control_: $\beta$ = `r round_x(s2_gam_ptab["factorbod_cog", "Estimate"], 2)`, $p$ `r ifelse(s2_gam_ptab["factorbod_cog", "Pr(>|t|)"] < 0.001, "< 0.001", paste("=", round_x(s2_gam_ptab["factorbod_cog", "Pr(>|t|)"], 3)))`) and a significant but small increase in capacities for _bodily sensation_ over the target age range (average smoothing term: $F$ = `r round_x(s2_gam_stab["s(target_year):factorBodily sensation", "F"], 2)`, $p$ `r ifelse(s2_gam_stab["s(target_year):factorBodily sensation", "p-value"] < 0.001, "< 0.001", paste("=", round_x(s2_gam_stab["s(target_year):factorBodily sensation", "p"], 3)))`, $edf$ = `r round_x(s2_gam_stab["s(target_year):factorBodily sensation", "edf"], 2)`).

Visual inspection of Figure 2, Panel A, clearly illustrates that the shape of perceived growth in children’s capacities varied substantially across domains. In the domain of _bodily sensation_, participants perceived hardly any growth over the target age range, and what growth they did perceive occurred in the first year of life. Participants perceived development in the domains that we have labeled _negative affect_ and _social connection_ to be quite non-linear, characterized by rapid development over the first 1-2 years of life, at which point the curves level off dramatically. In contrast, participants perceived dramatic development in the domain of _cognition and control_, but perceived this development to unfold fairly steadily across the full age range.

We speculate that these differences in perceptions of growth in children’s capacities emerge from intuitive theories of development that specify different mechanisms or drivers of development for different domains. For example, if participants believe that development in the domain of _bodily sensation_ is primarily driven by innate biological forces, they might assume that most of this development is complete before birth, leaving little room for change over the target ages assessed. [See @berent.etal2019 for empirical evidence in line with this speculation.] In contrast, in a domain where participants believe development depends on observation, exploration, or social learning, they might think that children need time in the world to change and grow, yielding lower estimates of capacities at birth and slower estimates of rates of change; or, in a domain where participants believe that development occurs through explicitly teaching the child something, they might perceive that development proceeds nonlinearly, such that large changes occur following specific milestones (e.g., entry to preschool). We explore these possibilities in Study 3.

```{r fig02, fig.cap="Perceived developmental trajectories for four domains of mental life among US adults (Studies 2-3). Lighter lines represent individual participants’ responses, black points correspond to mean responses across the sample, error bars are bootstrapped 95% confidence intervals, and thick red lines are predictions from our generalized additive models (beta regressions). In Study 2 (Panel A), participants assessed 5 capacities within each domain, and assessed all capacities for a given target age before moving on to the next target age. In Study 3 (Panel B), participants assessed 2 capacities within each domain, and assessed a single capacity for all target ages before moving on to the next capacity."}
knitr::include_graphics("../outputs/fig02.pdf")
```

@@

```{r s3 gam results, include = F}
s3_gam_ptab <- summary(s3_results$gam$gam)$p.table
s3_gam_stab <- summary(s3_results$gam$gam)$s.table
```

>>>>

Indeed, caring for a young child involves making countless decisions under uncertainty: Why is my child crying, smiling, shouting, or biting? What do they want, and what do they need? Can they understand what I am saying? Do they love me? As in other domains of reasoning[@gerstenberg.tenenbaum2017; @gopnik.wellman1994; @wellman.gelman1992], people likely draw on folk beliefs and intuitive theories about how the mind develops to predict, explain, and respond to children’s behavior and development, with consequences for the well-being of children in their care.

These folk beliefs and intuitive theories are, to date, poorly understood by social scientists. A handful of recent studies have examined the extent to which cognitive abilities, knowledge, and emotions are thought to be innate vs. learned, with special attention to comparing laypeople’s beliefs about innateness to scientific evidence for or against innateness and universality [@wang.feigenson2019; @berent.etal2019; @berent.etal2020; see @berent2021 for discussion]. These studies offer a fascinating glimpse of potential gaps, misconceptions, and biases in laypeople’s beliefs about the human mind, but they leave open many questions about the nature of the folk beliefs and intuitive theories themselves. What is the cognitive architecture underlying laypeople’s beliefs about the origins of the human mind? How do laypeople “carve up” the various sensations, experiences, and abilities that make up human mental life, and—most crucially from the perspective of caregiving—how do laypeople understand changes in mental life across early development?

In this paper, we examine how US adults conceptualize the developing minds of infants and children by drawing on techniques developed in studies of mind perception, in which researchers use large datasets and modern statistical methods to reveal conceptual representations of “the mind,” broadly construed. In a groundbreaking study, Gray et al. asked participants to compare the mental capacities of a range of humans, animals, technologies, and other entities, and argued that participants’ judgments revealed a fundamental distinction between “agency” (e.g., capacities for self-control, moral reasoning, memory, and planning) and “experience” (e.g., capacities for hunger, fear, pain, and pleasure) [@gray.etal2007]. More recently, Weisman et al. used a similar empirical approach to argue that people’s conceptions of mental life are anchored by three fundamental components: “body” (physiological sensations such as hunger and pain), “heart” (social-emotional abilities such as embarrassment and pride), and “mind” (perceptual-cognitive capacities such as vision and memory [@weisman.etal2017a; see @malle2019 for a similar three-factor solution]. Such findings have inspired subsequent work on topics as wide-ranging as moral reasoning, dehumanization and stereotyping, beliefs about the afterlife, and human-robot interactions.

@@

Yet infants and young children are far from unusual social partners. Caregiving relationships are at the core of human existence; for many people, infants and young children are among the most frequent and highly valued minds that we encounter. In the current studies we set out to shed new light on this important and understudied aspect of mind perception: how laypeople reason about the development of human mental life. In three studies, we assessed US adults’ folk beliefs about children’s developing abilities for physiological sensation, perception, cognition, emotion, self-regulation, social interaction, and other aspects of “mental life,” broadly construed. We focused on development between birth and age five years—a time of particularly dramatic maturation in perceptual acuity, emotional experience, cognitive ability, executive function, social engagement, and many other domains.

Methods, inclusion/exclusion criteria, and analyses marked as “preregistered” were preregistered on the Open Science Framework (OSF) website: (anonymized links for peer review: Study 1: https://osf.io/e6ajh/?view_only=f2539dbce3664172907aa7a381f62d34; Study 2: https://osf.io/j72dg/?view_only=7e4c01fec32f40a3b13c0ebc2b7ced18; Study 3: https://osf.io/xh8ce/?view_only=56649aff37aa4f1db2dc06b2e69576f4). All studies were approved by [IRB information removed to preserve anonymity]. For extended descriptions of methods and results, see Supplemental Materials.

@@

We designed Study 1 to explore the correlational structure of participants’ responses, as a window into the underlying conceptual structure that might support their reasoning about the development of mental life. Our primary analysis was a preregistered exploratory factor analysis (EFA) of participants’ capacity ratings. This analysis allowed us to examine which capacities tend to “hang together” in participants’ assessments of the mental lives of infants and young children. For example, when a participant indicated that newborns are highly capable of feeling scared, what other capacities did they tend to attribute to newborns? Following previous work on mind perception, we argue that the suites of capacities revealed by applying dimensionality reduction (in our case, EFA) to this covariance structure offer a meaningful approximation of the latent conceptual structure underlying participants’ reasoning about the developing human mind.

@@

It is worth noting that a handful of items loaded relatively strongly on two factors (loadings $\geq$ 0.40), suggesting that these factors are not entirely distinct. In our view, these cross-loadings were generally sensible. The item "feeling distressed", for instance, loaded roughly equally strongly on both _bodily sensation_ and _negative affect_—perhaps reflecting the idea that "distress" might be experienced both as a physical sensation (similar to pain) and an affective state (similar to frustration). Likewise, "seeing" loaded equally strongly on both _bodily sensation_ and _social connection_—perhaps reflecting an understanding of vision as both a basic sensory ability (closely tied to a particular body part) and a critical part of many social interactions (especially stereotypical interactions with infants, which often hinge on making eye contact, imitating facial expressions, playing peekaboo, etc.).

As a check on our choice of labels for these factors, we turned to OpenAI’s Generative Pre-trained Transformer [ChatGPT, version 4o; @openAI2024], an autoregressive transformer-based language prediction model trained on an enormous corpus of English-language text drawn primarily from “crawls” of the Internet as well as English-language novels and the open-source encyclopedia Wikipedia. Like other generative language models, ChatGPT was developed under the mandate of predicting the continuation of natural language sequences: Most concretely, then, its “responses” should be understood to reflect a sequence of words that is likely to follow the prompt it has been provided. In a more abstract sense, however, the content of these responses could be understood to represent the “common knowledge” embedded in the training data [for extended arguments relevant to this proposal, see @bhatia.richie2021; @lake.murphy2021; @petroni.etal2019]. For example, if prompted with the questions “What do you get when you mix red and yellow?” or “What’s a good name for a vegetarian restaurant?,” ChatGPT will generate text like “When you mix red and yellow, you get orange” or “The Green Plate” (respectively). In continuing the language sequence, the model also provides novel content that reflects actual human knowledge, biases, associations, and intuitions.

Using ChatGPT version 4o (https://chatgpt.com/), we provided a variety of prompts to the model to assess its “intuitions” about how to label the four categories of mental capacities suggested by the factor solution just described. One representative prompt began, “You're a statistician and a psychologist, and I'd like your help interpreting my exploratory factor analysis results,” followed by the EFA output taken directly from our analysis in R (containing the full list of factor loadings, as well as other summary characteristics of the EFA solution). The immediate response generated by ChatGPT included comments on model fit, factor structure, factor interpretability, factor contributions, chi-squared and BIC, as well as suggestions for “next steps”; under interpretability, it read, “MR1 [the factor we have referred to as _cognition and control_, although this label was never provided to the model] has high loadings on items related to planning, self-control, and goal-oriented behaviors. It could represent Self-Regulation or Executive Function. MR3 [what we have called _social connection_] includes items related to emotions such as feeling excited, happy, or loved. This could reflect Positive Emotional Experiences. MR2 [*bodily sensation*] has high loadings on items like feeling hungry, tired, and uncomfortable. This factor might capture Physical States or Needs. MR4 [*negative affect*] captures items like feeling helpless, frustrated, and annoyed. It may reflect Negative Emotional Responses or Distress.” When asked to tailor labels for an audience of cognitive scientists and clinical psychologists, the model suggested names like “Self-Regulation and Cognitive Control,” “Positive Affect and Social Emotions,” “Physiological Awareness and Basic Needs,” and “Negative Affect and Emotional Distress.” (See https://chatgpt.com/share/675a0c61-d534-8006-a978-b731bc997572 for the full text of these prompts and ChatGPT’s responses; OpenAI, 2024.) The similarity between our original labels and those generated by ChatGPT provides some evidence that the labels used are consistent with common knowledge and intuitions reflected in the vast swath of English language text that constituted the training data for this model, and by extension with the common knowledge and intuitions shared by many lay people in the English-speaking world. Nonetheless, we caution the reader to treat these labels as interpretations of empirical results, rather than empirical results themselves.

@@

@@

A preregistered EFA yielded four factors very similar to the four factors from Study 1, with all capacity items loading strongly on the factors that they were selected to represent and less strongly on other factors. (See Supplemental Materials for this 4-factor solution, which was suggested by both parallel analysis and Weisman et al.’s (2017) factor retention criteria, as well as the 8-factor solution suggested by minimizing BIC). Minimally, this suggests that our selection of mental capacities for this study captured the essential meaning of the four factors revealed in Study 1 (including the factor we have labeled _negative affect_: In Study 2, this factor accounted for fully `r round_x(s2_results$efa$Vaccounted["Proportion Explained", "MR1"]*100, 0)`% of the shared variance and had strong factor loadings ($\geq$ 0.60) for all 5 of the items selected to represent it, none of which loaded strongly on other factors). Beyond this, we consider these results to be a somewhat independent conceptual replication of Study 1 results and validation of our interpretation of these factors. After all, given that Study 2 featured only 20 of the 60 capacities featured in Study 1, it certainly could have been the case that this analysis would suggest retaining fewer factors (as often occurs with fewer variables). Conversely, given that Study 2 featured 13 within-subjects observations for each capacity compared to the 3 in Study 1, this analysis could have surfaced more than four factors (as often occurs with more observations). Finally, if we had misinterpreted the fundamental semantic content that distinguished the factors in Study 1 and selected inappropriate capacities to represent these factors, the Study 2 analysis could have identified more capacities with cross-loadings. Instead, the results of this analysis offer some validation of the four factors we have called _bodily sensation_, _negative affect_, _social connection_, and _cognition and control_.

@@

```{r s2 gam results, include = F}
s2_gam_ptab <- summary(s2_results$gam)$p.table
s2_gam_stab <- summary(s2_results$gam)$s.table
s2_gam_parameters <- parameters(s2_results$gam)
```

Regressing participants' item-level responses onto target age, domain, and interactions between them via a multilevel generalized additive model (using a binomial distribution and the logit link function) confirmed both of our preregistered hypotheses: On average, participants rated older children as more capable, but perceptions of the development of children’s capacities differed dramatically across domains. On one extreme, participants attributed very little capacity for _cognition and control_ to newborns (median score for newborns: `r s2_cap_rat_newborns_ss %>% filter(factor == "Cognition and control") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Cognition and control") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Cognition and control") %>% pull(sd) %>% round_x(., 2)`), but reported that such capacities increase dramatically and non-linearly over the target age range--that is, including a smoothing term for target age (analogous to a slope in a linear regression) significantly reduced the deviance of the model ($\chi^2$ = `r s2_gam_parameters %>% filter(Parameter == "s(target_year):factorCognition and control") %>% pull(z) %>% round_x(., 2)`, $p$ `r ifelse(s2_gam_parameters %>% filter(Parameter == "s(target_year):factorCognition and control") %>% pull(p)  < 0.001, "< 0.001", paste("=", round_x(s2_gam_parameters %>% filter(Parameter == "s(target_year):factorCognition and control") %>% pull(p), 3)))`); and the fitted smooth had moderate non-linear complexity (estimated degrees of freedom $edf$ = `r s2_gam_parameters %>% filter(Parameter == "s(target_year):factorCognition and control") %>% pull(df) %>% round_x(., 2)`). Compared to these attributions of _cognition and control_, participants attributed substantially greater capacities to newborns in the domains that we have labeled _social connection_ (median score for newborns: `r s2_cap_rat_newborns_ss %>% filter(factor == "Social connection") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Social connection") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Social connection") %>% pull(sd) %>% round_x(., 2)`; increase in log-odds relative to _cognition and control_: $\beta$ = `r round_x(s2_gam_ptab["factorsoc_cog", "Estimate"], 2)`, $p$ `r ifelse(s2_gam_ptab["factorsoc_cog", "Pr(>|z|)"] < 0.001, "< 0.001", paste("=", round_x(s2_gam_ptab["factorsoc_cog", "Pr(>|z|)"], 3)))`), but still perceived dramatic non-linear growth in this domain over the target age range ($\chi^2$ = `r s2_gam_parameters %>% filter(Parameter == "s(target_year):factorSocial connection") %>% pull(z) %>% round_x(., 2)`, $p$ `r ifelse(s2_gam_parameters %>% filter(Parameter == "s(target_year):factorSocial connection") %>% pull(p)  < 0.001, "< 0.001", paste("=", round_x(s2_gam_parameters %>% filter(Parameter == "s(target_year):factorSocial connection") %>% pull(p), 3)))`), $edf$ = `r s2_gam_parameters %>% filter(Parameter == "s(target_year):factorSocial connection") %>% pull(df) %>% round_x(., 2)`). In the domain of _negative affect_, participants attributed even greater capacities to newborns (median score for newborns: `r s2_cap_rat_newborns_ss %>% filter(factor == "Negative affect") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Negative affect") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Negative affect") %>% pull(sd) %>% round_x(., 2)`; increase in log-odds relative to _cognition and control_: $\beta$ = `r round_x(s2_gam_ptab["factorneg_cog", "Estimate"], 2)`, $p$ `r ifelse(s2_gam_ptab["factorneg_cog", "Pr(>|z|)"] < 0.001, "< 0.001", paste("=", round_x(s2_gam_ptab["factorneg_cog", "Pr(>|z|)"], 3)))`) and smaller increases over the target age range ($\chi^2$ = `r s2_gam_parameters %>% filter(Parameter == "s(target_year):factorSocial connection") %>% pull(z) %>% round_x(., 2)`, $p$ `r ifelse(s2_gam_parameters %>% filter(Parameter == "s(target_year):factorNegative affect") %>% pull(p)  < 0.001, "< 0.001", paste("=", round_x(s2_gam_parameters %>% filter(Parameter == "s(target_year):factorSocial connection") %>% pull(p), 3)))`), $edf$ = `r s2_gam_parameters %>% filter(Parameter == "s(target_year):factorSocial connection") %>% pull(df) %>% round_x(., 2)`). At the other extreme, in the domain of _bodily sensation_ participants attributed nearly maximal capacity to newborns (median score for newborns: `r s2_cap_rat_newborns_ss %>% filter(factor == "Bodily sensation") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Bodily sensation") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Bodily sensation") %>% pull(sd) %>% round_x(., 2)`; increase in log-odds relative to _cognition and control_: $\beta$ = `r round_x(s2_gam_ptab["factorbod_cog", "Estimate"], 2)`, $p$ `r ifelse(s2_gam_ptab["factorbod_cog", "Pr(>|z|)"] < 0.001, "< 0.001", paste("=", round_x(s2_gam_ptab["factorbod_cog", "Pr(>|z|)"], 3)))`) and the smallest increase increase in capacities over the target age range ($\chi^2$ = `r s2_gam_parameters %>% filter(Parameter == "s(target_year):factorBodily sensation") %>% pull(z) %>% round_x(., 2)`, $p$ `r ifelse(s2_gam_parameters %>% filter(Parameter == "s(target_year):factorNegative affect") %>% pull(p)  < 0.001, "< 0.001", paste("=", round_x(s2_gam_parameters %>% filter(Parameter == "s(target_year):factorBodily sensation") %>% pull(p), 3)))`), $edf$ = `r s2_gam_parameters %>% filter(Parameter == "s(target_year):factorBodily sensation") %>% pull(df) %>% round_x(., 2)`).

Modeling the effects of target age via smoothing terms allowed us to reconstruct the shape these perceived developmental trajectories from the bottom up, without specifying in advance the shape of the curve (but see Supplemental Materials for traditional logistic regressions). One consequence of this analysis choice is the model output does not include formal statistical tests of differences in perceived trajectories across domains—but visual inspection of Figure 2, Panel A, clearly illustrates that the shape of perceived growth in children’s capacities varied substantially across domains. In the domain of _bodily sensation_, participants perceived hardly any growth over the target age range, and what growth they did perceive occurred in the first year of life. Participants perceived development in the domains that we have labeled _negative affect_ and _social connection_ to be quite non-linear, characterized by rapid development over the first 1-2 years of life, at which point the curves level off dramatically. In contrast, participants perceived dramatic development in the domain of _cognition and control_, but perceived this development to unfold at a relatively steadier rate across the full age range.

We speculate that these differences in perceptions of growth in children’s capacities emerge from intuitive theories of development that specify different mechanisms or drivers of development for different domains. For example, if participants believe that development in the domain of _bodily sensation_ is primarily driven by innate biological forces, they might assume that most of this development is complete before birth, leaving little room for change over the target ages assessed. [See @berent.etal2019, for empirical evidence in line with this speculation.] In contrast, in a domain where participants believe development depends on observation, exploration, or social learning, they might think that children need time in the world to change and grow, yielding lower estimates of capacities at birth and slower estimates of rates of change; or, in a domain where participants believe that development occurs through explicitly teaching the child something, they might perceive that development proceeds nonlinearly, such that large changes occur following specific milestones (e.g., entry to preschool). We explore these possibilities in Study 3.

```{r fig02, fig.cap="Perceived developmental trajectories for four domains of mental life among US adults (Studies 2-3). Lighter lines represent individual participants’ responses, black points correspond to mean responses across the sample, error bars are bootstrapped 95% confidence intervals, and thick red lines are predictions from our multilevel generalized additive models. In Study 2 (Panel A), participants assessed 5 capacities within each domain, and assessed all capacities for a given target age before moving on to the next target age. In Study 3 (Panel B), participants assessed 2 capacities within each domain, and assessed a single capacity for all target ages before moving on to the next capacity."}
knitr::include_graphics("../outputs/fig02.pdf")
```

@@

```{r s3 gam results, include = F}
s3_gam_ptab <- summary(s3_results$gam)$p.table
s3_gam_stab <- summary(s3_results$gam)$s.table
s3_gam_parameters <- parameters(s3_results$gam)
```
